Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-01 draft/paper.tex
paper

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\Sg}{\Sigma}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\lamK}{\mathsf{lam}^{\K}}
\newcommand{\appK}{\mathsf{app}^{\K}}

@@

\newcommand{\bCon}{\boldsymbol{\Con}}
\newcommand{\bTy}{\boldsymbol{\Ty}}
\newcommand{\bSub}{\boldsymbol{\Sub}}
\newcommand{\bTm}{\boldsymbol{\Tm}}
\newcommand{\bGamma}{\boldsymbol{\Gamma}}
\newcommand{\bDelta}{\boldsymbol{\Delta}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\bt}{\boldsymbol{t}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bid}{\boldsymbol{\id}}
\newcommand{\bemptycon}{\boldsymbol{\emptycon}}
\newcommand{\bSet}{\boldsymbol{\Set}}
\newcommand{\bU}{\boldsymbol{\U}}
\newcommand{\bEl}{\boldsymbol{\El}}
\newcommand{\bPii}{\boldsymbol{\Pi}}
\newcommand{\bPie}{\boldsymbol{\Pie}}
\newcommand{\bPiinf}{\boldsymbol{\Piinf}}
\newcommand{\bappi}{\boldsymbol{\mathsf{app}}}
\newcommand{\blami}{\boldsymbol{\mathsf{lam}}}
\newcommand{\bId}{\boldsymbol{\Id}}

@@

\begin{definition}[Notion of signature]
A QIIT signature at level $i$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_i :\equiv (j : \mathbb{N})\ra(M : \ToS_{i,j})\ra Con_M
\]

@@

\subsection{Overview}

@@

In the following, we fix $i$ and $j$ levels, and define a model $M : \ToS_{i,
  1+\max(i,1+j)}$, such than $\Con_M$ is a type of structured categories (of
algebras). The level $i$ marks the level of all external sets in function
domains, and $j$ marks the level of all internal sorts in an algebra. Hence,
every algebra has level $\max(i,1+j)$. The bump is only needed for $j$, since
algebras only contain elements of $T : \Set_j$ types, while inductive sets are
themselves elements of $\Set_i$. For example, $NatAlg_i : \Set_{\max(0, 1+i)}$.

We present the components of the model in order. In the following, we use
\textbf{bold} font to disambiguate components of this model from internal
structures. For example, we use $\boldsymbol{\sigma : \Sub\,\Gamma\,\Delta}$
to denote a substitution in the model.

The model involves a large amount of technical detail, most of which we omit,
and only present the most salient parts.

@@

\begin{definition}[Finite limit CwFs]\label{def:flCwF}
For a level $i$, there is a cETT type $\flCwF_i : \Set_{1+i}$, which is an
iterated $\Sigma$-type with the following components:
\begin{enumerate}
  \item A CwF with underlying sets all in $\Set_i$. We reuse the component
    notations from Definition \ref{def:tos}.
  \item $\Sigma$-type $\Sg : (A : \Ty\,\Gamma)\ra \Ty\,(\Gamma \ext A)
    \ra \Ty\,\Gamma$, with term formers $\proj1$, $\proj2$ and $\blank,\blank$.
  \item Identity type $\Id : (A : \Ty\,\Gamma)\ra \Tm\,\Gamma\,A\ra
    \Tm\,\Gamma\,A\ra \Ty\,\Gamma$, with $\refl$, equality reflection and UIP.
  \item Constant families. This includes a type former $\Kfam : \Con \ra
    \Ty\,\Gamma$, where $\Gamma$ is implicitly quantified, together with $\lamK
    : \Sub\,\Gamma\,\Delta \ra \Tm\,\Gamma\,(\Kfam\,\Delta)$ and its inverse
    $\appK$. The idea is that $\Kfam\,\Delta$ is a representation of $\Delta$ as
    a type in any context. Clairambault and Dybjer called constant families
    ``democracy'' in \cite{clairambault2014biequivalence}.
\end{enumerate}
\end{definition}

@@

\begin{definition}[Notion of induction in an flCwF]
Given $\bGamma : \flCwF_i$, we have the following predicate on contexts:
\begin{alignat*}{3}
  & Inductive : \Con_{\bGamma} \ra \Set_i \\
  & Inductive\,\Gamma :\equiv (A : \Ty_{\bGamma}\,\Gamma)\ra \Tm_{\bGamma}\,\Gamma\,A
\end{alignat*}
\end{definition}

@@

@@

\subsection{Identity}
For $\bt$ and $\bu$ in $\bTm\,\bGamma\,(\bEl\,\ba)$, we define $\bId\,\bt\,\bu
\boldsymbol{:} \bTm\,\bGamma\,\bU$ as expressing pointwise equality of weak
sections.
\begin{alignat*}{3}
& (\bId\,\bt\,\bu)\,\ulGamma &&:\equiv \bt\,\ulGamma = \bu\,\ulGamma\\
& (\bId\,\bt\,\bu)\,\ulA     && :\equiv \lambda\,e.\, \bt\,\ulA = \bu\,\ulA
\end{alignat*}
Above, $\bt\,\ulA = \bu\,\ulA$ is well-typed because of $e :
\bt\,\ulGamma = \bu\,\ulGamma$. For substitutions, we have to complete a square
of equalities:
\begin{alignat*}{3}
  (\bId\,\bt\,\bu)\,(\ulsigma : \Sub\,\ulGamma\,\ulDelta) : \bt\,\ulGamma = \bt\,\ulGamma \ra
       \bt\,\ulDelta = \bt\,\ulDelta
\end{alignat*}
This can be given by $\bt\,\ulsigma : \ba\,\ulsigma\,(\bt\,\ulGamma) =
\bt\,\ulDelta$ and $\bu\,\ulsigma : \ba\,\ulsigma\,(\bu\,\ulGamma) =
\bu\,\ulDelta$. The action on terms is analogous. We omit preservation morphisms
here as they are straightforward. Like $\bPiinf$, $\bId$ also does not support
strict preservation of $\emptycon$ and $\ext$. Equality reflection and $\boldsymbol{\refl :}
\bId\,\bt\,\bt$ are also evident.

@@

>>>>

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\ToSSig}{\mathsf{ToSSig}}

\newcommand{\Sg}{\Sigma}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\lamK}{\mathsf{lam}^{\K}}
\newcommand{\appK}{\mathsf{app}^{\K}}

@@

\newcommand{\bCon}{\boldsymbol{\Con}}
\newcommand{\bTy}{\boldsymbol{\Ty}}
\newcommand{\bSub}{\boldsymbol{\Sub}}
\newcommand{\bTm}{\boldsymbol{\Tm}}
\newcommand{\bGamma}{\boldsymbol{\Gamma}}
\newcommand{\bDelta}{\boldsymbol{\Delta}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\bt}{\boldsymbol{t}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bid}{\boldsymbol{\id}}
\newcommand{\bemptycon}{\boldsymbol{\emptycon}}
\newcommand{\bSet}{\boldsymbol{\Set}}
\newcommand{\bU}{\boldsymbol{\U}}
\newcommand{\bEl}{\boldsymbol{\El}}
\newcommand{\bPii}{\boldsymbol{\Pi}}
\newcommand{\bPie}{\boldsymbol{\Pie}}
\newcommand{\bPiinf}{\boldsymbol{\Piinf}}
\newcommand{\bappi}{\boldsymbol{\mathsf{app}}}
\newcommand{\blami}{\boldsymbol{\mathsf{lam}}}
\newcommand{\bId}{\boldsymbol{\Id}}
\newcommand{\bM}{\boldsymbol{\mathsf{M}}}

@@

\begin{definition}[Notion of signature]\label{def:signature}
A QIIT signature at level $i$ is a context in an arbitrary p$M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_i :\equiv (j : \mathbb{N})\ra(M : \ToS_{i,j})\ra \Con_M
\]

@@

\subsection{Overview}\label{sec:overview}

@@

In the following, for $i$ and $j$ levels we define a model $\bM_{i,j} : \ToS_{i,
  1+\max(i,1+j)}$, such than $\Con_{\bM_{i,j}}$ is a type of structured categories (of
algebras). The level $i$ marks the level of all external sets in function
domains, and $j$ marks the level of all internal sorts in an algebra. Hence,
every algebra has level $\max(i,1+j)$. The bump is only needed for $j$, since
algebras only contain elements of $T : \Set_j$ types, while inductive sets are
themselves elements of $\Set_i$. For example, $NatAlg_i : \Set_{\max(0, 1+i)} :
\Set_{1+\max(0, 1+i)}$,

We present the components of the model in order. In the following, we use
\textbf{bold} font to disambiguate components of $\bM_{i,j}$ from components of other
structures. For example, we use $\boldsymbol{\sigma : \Sub\,\Gamma\,\Delta}$ to
denote a substitution in $\bM_{i,j}$.

The model involves a large amount of technical detail; we omit a significant
part of this, and only present the most salient parts.

@@

\begin{definition}[Finite limit CwFs]\label{def:flCwF}
For each level $i$ we define $\flCwF_i : \Set_{1+i}$ as an iterated $\Sigma$-type
with the following components:
\begin{enumerate}
  \item A CwF with underlying sets all in $\Set_i$. We reuse the component
    notations from Definition \ref{def:tos}.
  \item $\Sigma$-type $\Sg : (A : \Ty\,\Gamma)\ra \Ty\,(\Gamma \ext A)
    \ra \Ty\,\Gamma$, with term formers $\proj1$, $\proj2$ and $\blank,\blank$.
  \item Identity type $\Id : (A : \Ty\,\Gamma)\ra \Tm\,\Gamma\,A\ra
    \Tm\,\Gamma\,A\ra \Ty\,\Gamma$, with $\refl$, equality reflection and UIP.
  \item Constant families. This includes a type former $\Kfam : \Con \ra
    \Ty\,\Gamma$, where $\Gamma$ is implicitly quantified, together with $\lamK
    : \Sub\,\Gamma\,\Delta \ra \Tm\,\Gamma\,(\Kfam\,\Delta)$ and its inverse
    $\appK$. The idea is that $\Kfam\,\Delta$ is a representation of $\Delta$ as
    a type in any context. Clairambault and Dybjer called constant families
    ``democracy'' in \cite{clairambault2014biequivalence}.
\end{enumerate}
\end{definition}

@@

\begin{definition}[Notion of induction in an flCwF]\label{def:induction}
Given $\bGamma : \flCwF_i$, we have the following predicate on contexts:
\begin{alignat*}{3}
  & Inductive : \Con_{\bGamma} \ra \Set_i \\
  & Inductive\,\Gamma :\equiv (A : \Ty_{\bGamma}\,\Gamma)\ra \Tm_{\bGamma}\,\Gamma\,A
\end{alignat*}
\end{definition}

@@

$\boldsymbol{\appi : \Tm\,\Gamma\,(\Pii\,a\,B)\ra
  \Tm\,(\Gamma\,\ext\,\El\,a)\,B}$ can be defined as currying of the underlying
functions, and $\boldsymbol{\lami}$ as uncurrying.

@@

\subsection{Identity}
For $\bt$ and $\bu$ in $\bTm\,\bGamma\,(\bEl\,\ba)$, we define $\bId\,\bt\,\bu
\boldsymbol{:} \bTm\,\bGamma\,\bU$ as expressing pointwise equality of weak
sections.
\begin{alignat*}{3}
& (\bId\,\bt\,\bu)\,\ulGamma &&:\equiv \bt\,\ulGamma = \bu\,\ulGamma\\
& (\bId\,\bt\,\bu)\,\ulA     && :\equiv \lambda\,e.\, \bt\,\ulA = \bu\,\ulA
\end{alignat*}
Above, $\bt\,\ulA = \bu\,\ulA$ is well-typed because of $e :
\bt\,\ulGamma = \bu\,\ulGamma$. For substitutions, we have to complete a square
of equalities:
\begin{alignat*}{3}
  (\bId\,\bt\,\bu)\,(\ulsigma : \Sub\,\ulGamma\,\ulDelta) : \bt\,\ulGamma = \bu\,\ulGamma \ra
       \bt\,\ulDelta = \bu\,\ulDelta
\end{alignat*}
This can be given by $\bt\,\ulsigma : \ba\,\ulsigma\,(\bt\,\ulGamma) =
\bt\,\ulDelta$ and $\bu\,\ulsigma : \ba\,\ulsigma\,(\bu\,\ulGamma) =
\bu\,\ulDelta$. The action on terms is analogous. We omit preservation morphisms
here as they are straightforward. Like $\bPiinf$, $\bId$ also does not support
strict preservation of $\emptycon$ and $\ext$. Equality reflection and $\boldsymbol{\refl :}
\bId\,\bt\,\bt$ are also evident.

With this, we have defined the $\bM_{i,j} : \ToS_{i, 1+\max(i,1+j)}$ model that we set out to define
in Section \ref{sec:overview}.

@@

At this point, we only have a notion of algebra for ToS, from Definition
\ref{def:tos}. In the following sections, we would also like to talk about
initial ToS-algebras and ToS-induction. We get these notions by giving a QIIT
signature for ToS, and interpreting it in the $\bM$ model from the previous
section.

\begin{definition}[Signature for ToS] For each level $i$,
we define $\ToSSig_i : \Sig_{1+i}$, as the signature for the theory of
signatures with external sets in $\Set_i$. This is a large and infinitary QIIT
signature, as we have $\Pie$ and $\Piinf$ abstracting over $A : \Set_i$ and
branching with $A \ra \Ty\,\Gamma$ and $A \ra \Tm\,\Gamma\,\U$ respectively. We
present an excerpt from $\ToSSig_i$ below.
\begin{alignat*}{3}
  & \emptycon && \ext (Con : \U)\\
  &           && \ext (Sub : Con\arri Con \arri \U)\\
  &           && \ext (Ty  : Con\arri\U)\\
  &           && \ext (Tm  : \Pii(\Gamma : Con)(\Ty\,\Gamma \arri \U))\\
  &           && ...\\
  &           && \ext (\Piinf : \Pii(\Gamma : Con)\\
  &           && \hspace{4.2em}(\Pie \Set_i (\lambda\,A.\,(A \arrinf \Ty\,\Gamma)\arri\El(\Ty\,\Gamma))))\\
  &           && ...
\end{alignat*}

Now, for each $j$, the interpretation of $\ToSSig_i$ in $\bM_{1+i,j}$ yields an
flCwF $\bGamma$ such that $\Con_{\bGamma} = \ToS_{i,j}$. In short, we can
recover ToS algebras from $\ToSSig$. This follows by computation of the
interpretation and the fact that $\ToSSig$ is the internal representation of
$\ToS$. Also, since we have an flCwF of $\ToS_{i,j}$-algebras, we can use
Definition \ref{def:induction} for the notion of $\ToS$-induction.

\emph{Remark.} By the definition of $\bemptycon$ and
$\boldsymbol{\blank\ext\blank}$, the types of algebras computed by $\bM$ are
always left-nested iterated $\Sigma$-types which start with $\top$. Hence, we
need to require that Definition \ref{def:tos} is similarly left-nested and
starts with $\top$, in order to make the match strict.

\section{Term Models of QIITs}

%% \begin{theorem}  For each $j$, the interpretation of $\ToSSig_i$ in $\bM_{1+i,j}$
%% yields an flCwF $\bGamma$ such that $\Con_{\bGamma} = \ToS_{i,j}$. In short, we
%% can recover ToS algebras from $\ToSSig$.
%% \end{theorem}
%% \begin{proof}
%% \end{proof}

%% Overall, this is just rephrasing Definition \ref{def:tos}
%% internally to an arbitrary $M : \ToS_{1+i,j}$; recall from Definition
%% \ref{def:signature} that $\Sig_{1+i} = (j : \mathbb{N})(M : \ToS_{1+i,j})\ra
%% p\Con_M$. For example, the underlying sets are specified as
